---
title: "The Fuel Consumption of Automatic and Manual Vehicles (1973-74 models)"
author: "Jonathan Hill, MPA"
date: "Thursday, April 23, 2015"
output: html_document
---
```{r,warning=FALSE}
library(datasets)
library(lattice)
data(mtcars)
```

# Overview

The two questions addressed in this project are:

1. Is an automatic or manual transmission better for MPG?
2. What is the MPG difference between automatic and manual transmissions?

I will attempt to answer these questions using a multi-linear regression model with a dependent variable of miles per gallon.  The model will include independ variables (control variables) for the characteristics of the engine, such as the number of cylinders and displacement volume, and the car itself, such as rear axle ratio and weight.

## The Data
The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and 10 aspects of automobile design and performance for 32 automobiles (1973-74 models).

Variable Name  | Description
------------- | -------------
*mpg* | Miles/(US) gallon
**am**  | Transmission (0 = automatic, 1 = manual)
cyl  | Number of cylinders
disp  | Displacement (cu.in.)
hp  | Gross horsepower
drat  | Rear axle ratio
wt  | Weight (lb/1000)
qsec  | 1/4 mile time
vs  | V/S
gear  | Number of forward gears
carb  | Number of carburetors


# Exploratory Analysis

**Basic Summary Statistics**
```{r,echo=FALSE}
summary(mtcars)
```
```{r,echo=FALSE}
hist(mtcars$mpg,breaks=32,col="red",main="Histogram of Dependent Variable (mpg)",xlab="Miles per Gallon")
abline(v=mean(mtcars$mpg),lwd=2)
shapiro.test(mtcars$mpg)
```

The result of the Shapiro-Wilk normality test suggests it is safe to assume the dependent variable is normal with a mean of `r round(mean(mtcars$mpg),2)` mpg (as illustrated by the dark black line through the histogram above).

```{r,echo=FALSE}
mtcars$am <- as.factor(mtcars$am)
levels(mtcars$am) <- c("Automatic","Manual")
mtcars$cyl <- as.factor(mtcars$cyl)
levels(mtcars$cyl) <- c("4 cylinders","6 cylinders","8 cylinders")
xyplot(mpg~hp|cyl*am,data=mtcars,ylab="Miles per Gallon",xlab="Horsepower",main="Engine Characteristics and MPG")
```

```{r,echo=FALSE}
xyplot(mpg~wt|cyl*am,data=mtcars,ylab="Miles per Gallon",xlab="Weight (lb/1000)",main="Vehicle Weight and MPG")
```

These plots seem to suggest that manual transmission vehicles may be more fuel efficient than automatic transmission vehicles.  The top row appears to be slightly higher than the bottom row, but they also illustrate that other factors are definitely effecting fuel efficiency.

For example, there seems to be an indirect relationship between fuel efficiency and weight.  This builds confidence in the quality of our data because it is reasonable to assume that heavier cars are less fuel efficient.  It also necessitates the use of a multi-variate method such as multi-linear regression in order to accurately assess the impact that a car's transmission type (automatic/manual) effects its fuel efficiency.

